Model Search in Relational Domains

### 2016-10-19 algorithm.tex
notes from today's call

\section{Abstract Conflict Driven Clause Learning}
%
\begin{algorithm2e}[t]
\DontPrintSemicolon
\SetKwFunction{decide}{Decide}
\SetKwFunction{deduce}{deduce}
\SetKwFunction{backtrack}{backtrack}
\SetKwFunction{print}{print}
\SetKwFunction{return}{return}
\SetKwFunction{nondet}{nondet}
\SetKwFunction{run}{run}
\SetKwFunction{learn}{learn}
\SetKwFunction{continue}{continue}
\SetKwFunction{assign}{assign}
\SetKwRepeat{Do}{do}{while}
%\SetKwFunction{assume}{assume}
%\SetKwFunction{isf}{isFeasible}
\SetKwData{conflict}{CONFLICT}
\SetKwData{safe}{safe}
\SetKwData{unsafe}{unsafe}
\SetKwData{unknown}{unknown}
\SetKwData{true}{true}
\SetKwInOut{Input}{input}
\SetKwInOut{Output}{output}
\SetKwFor{Loop}{Loop}{}{}
\SetKw{KwNot}{not}
\begin{small}
\Input{Set $S$ of abstract deduction transformer, 
$\widehat{ded_{\varphi}}$ for a formula $\varphi$ over an abstract domain $A$}
\Output{The status (\safe or \unsafe) and a counterexample if \unsafe}
$v \leftarrow \top$ \;
$\mathcal{T} \leftarrow \epsilon$ \;
$\mathcal{R} \leftarrow \emptyset$ \;
$result \leftarrow \deduce(S,v,\mathcal{T},\mathcal{R})$ \;
\uIf{$result$ == \conflict} {
  \return \safe \;
}
\While{$true$} 
{
  \uIf{$v$ is $\gamma$-complete} {
    \return \unsafe;
  }
  $d \leftarrow decide(v)$ \;
  $v \leftarrow v \meet d$ \; 
  $\mathcal{T} \leftarrow \mathcal{T} . d$ \; 
  $\mathcal{R}[|\mathcal{T}|] \leftarrow nil$ \;
  $result \leftarrow \deduce(S,v,\mathcal{T},\mathcal{R})$\;
  \Do{$result == \conflict$} {
    learn() \;
    \uIf{$\neg$ conflict\_analysis()} {
      return \safe \;  
    }
    result $\leftarrow \deduce(S,v,\mathcal{T},\mathcal{R})$ \;
  }
}
\end{small}
\caption{Abstract Conflict Driven Clause Learning $\langle ACDCL(S) \rangle$ \label{Alg:acdcl}}
\end{algorithm2e}
%
%\subsection{Algorithm for Abstract Conflict Driven Clause Learning}
Algorithm~\ref{Alg:acdcl} presents an overview of the ACDCL algorithm.
The algorithm begins with a set $S$ of abstract deduction transformer 
$\widehat{ded_{\varphi}}$ \pscmt{???} for a formula $\varphi$ over an abstract 
domain $A$, an abstract valuation $\langle v \leftarrow \top \rangle$, 
and an empty propagation trail and reason trail, $\mathcal{T}$, $\mathcal{R}$, 
respectively.  The propagation trail stores all meet irreducibles inferred 
by the abstract model search phase (deductions and decisions), while the 
reason trail maps elements of propagation trail to transformers that were 
used to derive them.

The procedure $deduce(S,v,\mathcal{T},\mathcal{R})$ computes a greatest 
fixed point over the transformers in F \pscmt{what is F?} that refines the abstract valuation 
$v$, similar to the BCP \pscmt{mentioned before?} step in SAT solvers.  If the result of $deduce$ 
returns {\em CONFLICT} \pscmt{don't use capitals} ($\perp$), the algorithm terminates with {\em safe}. 
Else, the analysis enters into the while($true$) loop and makes a new decision 
by a call to $decide(v)$ which returns a new meet irreducible $d$ that is
consistent to the state of the solver. \pscmt{why do we have to store $v$ and $\mathcal{T}$?} If $d$ refines the current abstract 
value $v$, the corresponding propagation and reason trail, $\mathcal{T}$ 
and $\mathcal{R}$ are updated.

For example, a decision in interval domain restricts the range of 
intervals for variables, so the analysis jumps under a 
greatest fixed-point \pscmt{???}. Note that the widening operation in abstract interpretation 
jumps above a least fixed-point, so decisions can be viewed as 
{\em dual widening} \pscmt{what do we learn?}.  The procedure $deduce$ is called next to infer new 
set of meet irreducibles for the current decision.  The model search phase 
alternates between the decision and deduction until the abstract value $v$ 
precisely represents the set of models or the deduction leads to a {\em CONFLICT}.  
Checking whether $v$ precisely represents set of models corresponds to 
$\gamma$-completeness~\cite{dhk2013-popl} check in abstract interpretation.  
If $v$ is $\gamma$-complete, then it cannot be refined further.  Thus, the 
algorithm returns the error trace $v$ \pscmt{this is not a trace} and terminates with UNSAFE or the 
current abstraction is insufficient to determine the satisfiability of $\varphi$.  
However, in case of {\em CONFLICT}, the algorithm enters in the 
$conflict\_analysis()$ phase to learn the reason for {\em conflict}.  
There can be multiple incomparable reasons for conflict -- based on 
the choice of Unique Implication Point (UIP)~\cite{cdcl}.  ACDCL heuristically 
chooses one reason.  A learnt clause must include asserting cuts which 
guarantees derivation of new meet irreducibles after backtracking. The clause 
learning and backtracking continues as long as the result of deduction 
is conflicting ($\perp$), that is, the abstract valuation $v$ does not
abstractly satisfy the formula \pscmt{define}.  If no further backtrack is possible, 
then the algorithm terminates and $\varphi$ is {\em SAFE}. Else, the 
algorithm makes a new decision and the above process is repeated 
until a real counterexample \pscmt{is this a model?} is obtained or the algorithm backtracks to 
decision level 0 after a conflict in which case it returns {\em SAFE}.

>>>>

\section{Abstract Conflict Driven Clause Learning}
%
\begin{algorithm2e}[t]
\DontPrintSemicolon
\SetKwFunction{decide}{Decide}
\SetKwFunction{deduce}{deduce}
\SetKwFunction{backtrack}{backtrack}
\SetKwFunction{print}{print}
\SetKwFunction{return}{return}
\SetKwFunction{nondet}{nondet}
\SetKwFunction{run}{run}
\SetKwFunction{learn}{learn}
\SetKwFunction{continue}{continue}
\SetKwFunction{assign}{assign}
\SetKwRepeat{Do}{do}{while}
%\SetKwFunction{assume}{assume}
%\SetKwFunction{isf}{isFeasible}
\SetKwData{conflict}{CONFLICT}
\SetKwData{safe}{safe}
\SetKwData{unsafe}{unsafe}
\SetKwData{unknown}{unknown}
\SetKwData{true}{true}
\SetKwInOut{Input}{input}
\SetKwInOut{Output}{output}
\SetKwFor{Loop}{Loop}{}{}
\SetKw{KwNot}{not}
\begin{small}
\Input{Set $S$ of abstract deduction transformer, 
$\widehat{ded_{\varphi}}$ for a formula $\varphi$ over an abstract domain $A$}
\Output{The status (\safe or \unsafe) and a counterexample if \unsafe}
$v \leftarrow \top$ \;
$\mathcal{T} \leftarrow \epsilon$ \;
$\mathcal{R} \leftarrow \emptyset$ \;
$result \leftarrow \deduce(S,v,\mathcal{T},\mathcal{R})$ \;
\uIf{$result$ == \conflict} {
  \return \safe \;
}
\While{$true$} 
{
  \uIf{$v$ is $\gamma$-complete} {
    \return \unsafe;
  }
  $d \leftarrow decide(v)$ \;
  $v \leftarrow v \meet d$ \; 
  $\mathcal{T} \leftarrow \mathcal{T} . d$ \; 
  $\mathcal{R}[|\mathcal{T}|] \leftarrow nil$ \;
  $result \leftarrow \deduce(S,v,\mathcal{T},\mathcal{R})$\;
  \Do{$result == \conflict$} {
    learn() \;
    \uIf{$\neg$ conflict\_analysis()} {
      return \safe \;  
    }
    result $\leftarrow \deduce(S,v,\mathcal{T},\mathcal{R})$ \;
  }
}
\end{small}
\caption{Abstract Conflict Driven Clause Learning $\langle ACDCL(S) \rangle$ \label{Alg:acdcl}}
\end{algorithm2e}
%
%\subsection{Algorithm for Abstract Conflict Driven Clause Learning}
Algorithm~\ref{Alg:acdcl} presents an overview of the ACDCL algorithm.
The algorithm begins with a set $S$ of abstract deduction transformer 
$\widehat{ded_{\varphi}}$ \pscmt{define in terms of P and A} for a formula $\varphi$ over an abstract 
domain $A$, an abstract valuation $\langle v \leftarrow \top \rangle$, 
and an empty propagation trail and reason trail, $\mathcal{T}$, $\mathcal{R}$, 
respectively.  The propagation trail stores all meet irreducibles inferred 
by the abstract model search phase (deductions and decisions), while the 
reason trail maps elements of propagation trail to transformers that were 
used to derive them. \pscmt{define operator to turn trail into abstract value $\bigsqcap\mathcal{T}=\bigsqcap_{m\in\mathcal{T}}m$}

The procedure $deduce(S,v,\mathcal{T},\mathcal{R})$ computes a greatest 
fixed point over the transformers in $\varphi$ that refines the abstract valuation 
$v$, similar to the BCP step in SAT solvers.  If the result of $deduce$ 
returns {\em CONFLICT} \pscmt{don't use capitals} ($\perp$), the algorithm terminates with {\em safe}. 
Else, the analysis enters into the while($true$) loop and makes a new decision 
by a call to $decide(v)$ which returns a new meet irreducible $d$ that is
consistent to the state of the solver. \pscmt{why do we have to store $v$ and $\mathcal{T}$?} If $d$ refines the current abstract 
value $v$, the corresponding propagation and reason trail, $\mathcal{T}$ 
and $\mathcal{R}$ are updated.

For example, a decision in the interval domain restricts the range of 
intervals for variables.
% so the analysis jumps under a 
%greatest fixed-point \pscmt{???}. Note that the widening operation in abstract interpretation 
%jumps above a least fixed-point, so decisions can be viewed as 
%{\em dual widening} \pscmt{what do we learn?}.  
The procedure $deduce$ is called next to infer a new 
set of meet irreducibles for the current decision.  The model search phase 
alternates between the decision and deduction until the abstract value $v$ 
precisely \pscmt{not true} represents the set of models or the deduction leads to a {\em CONFLICT}.  
Checking whether the abstract model $v$ concretises to a set of models of $\varphi$; this corresponds to a 
$\gamma$-completeness~\cite{dhk2013-popl} check in abstract interpretation.  
If $v$ is $\gamma$-complete, then it cannot be refined further.  Thus, the 
algorithm returns the abstract model $v$, which is a set of concrete models, and terminates with UNSAFE or the 
current abstraction is insufficient to determine the satisfiability of $\varphi$.  
However, in case of {\em CONFLICT}, the algorithm enters in the 
$conflict\_analysis()$ phase to learn the reason for {\em conflict}.  
There can be multiple incomparable reasons for conflict -- based on 
the choice of Unique Implication Point (UIP)~\cite{cdcl}.  ACDCL heuristically 
chooses one reason.  A learnt clause must include asserting cuts which 
guarantees derivation of new meet irreducibles after backtracking. The clause 
learning and backtracking continues as long as the result of deduction 
is conflicting ($\perp$), that is, the abstract valuation $v$ does not
abstractly satisfy the formula \pscmt{define}.  If no further backtrack is possible, 
then the algorithm terminates and $\varphi$ is {\em SAFE}. Else, the 
algorithm makes a new decision and the above process is repeated 
until a real counterexample \pscmt{is this a model?} is obtained or the algorithm backtracks to 
decision level 0 after a conflict in which case it returns {\em SAFE}.

### 2016-10-19 deductions.tex
notes from today's call

Algorithm~\ref{Alg:ms} presents the deduction phase in abstract model search procedure.
A {\em worklist} is initialized following a heuristics choice for iteration 
strategies. A {\em forward} iteration strategy initializes the worklist with 
transformers that has constants in the right-hand side. Whereas, a {\em backward} 
iteration strategy starts with a set of assertions. On the other hand, a 
{\em chaotic} \pscmt{that's not a good name because it is not chaotic in the meaning of chaotic iterations} iteration strategy initializes the worklist with set of all 
transformers.  For every meet irreducible deduction transformer \pscmt{rename} $s \in S$ 
passed to the abstract domain, if the domain infers a new meet irreducible 
$\mathit{meet\_irrd}$, it is added to $\mathcal{T}$ \pscmt{It usually infers several ones.}.  Additionally, the reason 
trail $\mathcal{R}$ \pscmt{define} is updated with the transformer $s$ as the reason for inferring
$\mathit{meet\_irrd}$. If a conflict is deduced, that is, $\mathit{meet\_irrd} = \bot$ 
is inferred by the domain, the algorithm terminates with CONFLICT.  However,
when a fixed-point is reached, the abstract valuation $v$ is checked for set of
models.  This is determined by checking whether the set of assignments in $v$ is 
$\gamma$-complete~\cite{dhk2013-popl}.  This can be achieved by checking if there is a 
concrete solution $c$ such that the inequality $c(x) \subseteq \gamma \circ
v(x)$ holds for every meet irreducible $x$ \pscmt{No, you have to consider $\bigsqcap x$}.  If $v$ is $\gamma$-complete, that is, the 
abstract valuation $v$ is abstractly satisfying, the abstract model 
search terminates with SAT.  Otherwise, the algorithm returns UNKNOWN and the 
model search procedure makes a new decision.

>>>>

Algorithm~\ref{Alg:ms} presents the deduction phase in abstract model search procedure.
A {\em worklist} is initialized following a heuristics choice for iteration 
strategies. A {\em forward} iteration strategy initializes the worklist with 
transformers that has constants in the right-hand side. Whereas, a {\em backward} 
iteration strategy starts with a set of assertions. On the other hand, a 
{\em multi-way} iteration strategy initializes the worklist with set of all 
transformers.  For every meet irreducible deduction transformer \pscmt{rename} $s \in S$ 
passed to the abstract domain, if the domain infers a new meet irreducible 
$\mathit{meet\_irrd}$, it is added to $\mathcal{T}$ \pscmt{It usually infers several ones.}.  Additionally, the reason 
trail $\mathcal{R}$ \pscmt{define} is updated with the transformer $s$ as the reason for inferring
$\mathit{meet\_irrd}$. If a conflict is deduced, that is, $\mathit{meet\_irrd} = \bot$ 
is inferred by the domain, the algorithm terminates with CONFLICT.  However,
when a fixed-point is reached, the abstract valuation $v$ is checked for set of
models.  This is determined by checking whether the set of assignments in $v$ is 
$\gamma$-complete~\cite{dhk2013-popl}. If $v$ is $\gamma$-complete, that is, the 
abstract valuation $v$ is abstractly satisfying, the abstract model 
search terminates with SAT.  Otherwise, the algorithm returns UNKNOWN and the 
model search procedure makes a new decision.

### 2016-10-19 graphs.tex
notes from today's call

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{tabular}{@{\hspace{-1.5em}}c@{\hspace{1em}}c}
\begin{tikzpicture}[scale=0.75]
	\begin{loglogaxis} [xmin=0.1,xmax=2500, ymin=0.1, ymax=2500, xlabel=Forward
  Propagation, ylabel=Chaotic Propagation, 
			legend style={at={(0.8,0.15)},
			anchor=north,legend columns=-1 },
			]
\addplot [mark size=1pt,only marks,scatter,point meta=explicit symbolic,
	scatter/classes={a={mark=square,draw=black}}] 
	table [meta=label] {plotdata/scatter-chaotic-forward.dat};
	%\legend{Safe,Unsafe}
  \addplot [domain=.1:2500] {x};
\end{loglogaxis}
\end{tikzpicture}
 &
\begin{tikzpicture}[scale=0.75]
	\begin{axis} [
  title={Performance of Decision Heuristics},
  xmin=0, xmax=70, 
  ymin=0, ymax=1500, 
  xlabel={Benchmarks},
  ylabel={time (in seconds)},
  xtick={},
  ytick={}, %{0, 0.5, 1.0, 10, 50, 100, 200, 500, 700, 1000, 1500},
	legend pos=north west,
 ]
 \addplot table[x=benchmark,y=time] {plotdata/berkmin.dat};
 %\addplot table[x=benchmark,y=time] {plotdata/random.dat};
 \legend{Berkmin}
 %\legend{Random}
 %\addplot [domain=.1:2500] {x};
 \end{axis}
\end{tikzpicture} \\
(a) & (b)
\end{tabular}
\caption{\label{prop-dec}
Effect of Propagation Strategies and Decision Heuristics 
}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{tabular}{@{\hspace{-1.5em}}c@{\hspace{1em}}c}
\begin{tikzpicture}[scale=0.75]
	\begin{loglogaxis} [xmin=0.1,xmax=2500, ymin=0.1, ymax=2500, xlabel=Forward
  Propagation, ylabel=Chaotic Propagation, 
			legend style={at={(0.8,0.15)},
			anchor=north,legend columns=-1 },
			]
\addplot [mark size=1pt,only marks,scatter,point meta=explicit symbolic,
	scatter/classes={a={mark=square,draw=black}}] 
	table [meta=label] {plotdata/scatter-chaotic-forward.dat};
	%\legend{Safe,Unsafe}
  \addplot [domain=.1:2500] {x};
\end{loglogaxis}
\end{tikzpicture}
 &
\begin{tikzpicture}[scale=0.75]
	\begin{axis} [
  title={Performance of Decision Heuristics}, ymode = log,
  xmin=0, xmax=70, 
  ymin=0, ymax=1500, 
  xlabel={Benchmarks},
  ylabel={time (in seconds)},
  xtick={},
  ytick={}, %{0, 0.5, 1.0, 10, 50, 100, 200, 500, 700, 1000, 1500},
	legend pos=north west,
 ]
 \addplot table[x=benchmark,y=time] {plotdata/berkmin.dat};
 \addplot table[x=benchmark,y=time] {plotdata/random.dat};
 \addlegendentry{Berkmin}
 \addlegendentry{Random}
 %\addplot [domain=.1:2500] {};
 \end{axis}
\end{tikzpicture} \\
(a) & (b)
\end{tabular}
\caption{\label{prop-dec}
Effect of Propagation Strategies and Decision Heuristics 
}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2016-10-19 learning.tex
notes from today's call

\paragraph {\textbf{Lifting UIP to relational domains}}
An UIP is a special node in the abstract conflict graph such that 
any path from the last decision node to the conflict node must pass 
through it.  Unlike a SAT solver, a UIP in ACDCL may have meet irreducibles 
that involve the same variables.  The first-UIP is a node closest \pscmt{unique?} to the 
conflict, and the last UIP is the decision node itself.  Computing UIPs ensures 
asserting cuts, that is, it yields clauses that generates new deduction after 
backtracking.  Every cut in the graph reason for conflict that can be used in learning.  
A first-UIP algorithm traverses the trail $\mathcal{T}$ starting from the 
conflict node and iteratively finds new nodes that contradict \pscmt{we have to define what that means in the preliminaries} with the 
transformer stored in $reasons[\bot]$.

>>>>

\paragraph {\textbf{Lifting UIP to relational domains}}
An UIP is a special node in the abstract conflict graph such that 
any path from the last decision node to the conflict node must pass 
through it.  Unlike a SAT solver, a UIP in ACDCL may have meet irreducibles 
that involve the same variables.  The first-UIP is a node closest \pscmt{unique?} to the 
conflict, and the last UIP is the decision node itself.  Computing UIPs ensures 
asserting cuts, that is, it yields clauses that generates new deduction after 
backtracking.  Every cut in the graph reason for conflict that can be used in learning.  
A first-UIP algorithm traverses the trail $\mathcal{T}$ starting from the 
conflict node and iteratively finds new nodes that contradict \pscmt{we have to define what that means in the preliminaries} with the 
transformer stored in $\mathcal{R}[\bot]$.
